---
title: "map"
author: "Chuhan Yue"
date: "2023-10-27"
output: pdf_document
---

```{r setup, include=FALSE}
library(plotly)
library(tidyverse)
```

```{r}
data<-read.csv("eqarchive-en.csv")
head(data)
```

```{r}

fig <- data %>% plot_ly( 
    lat = ~latitude,
    lon = ~longitude,
    marker = list(color = ~magnitude, 
      colorscale = 'Viridis',
      size=3),
    type = 'scattermapbox', mode="markers",text = ~paste("Magnitude: ", magnitude)) 
fig <- fig %>%
  layout(
    mapbox = list(
      style = 'open-street-map',
      zoom =2.5,
      center = list(lon = -88, lat = 34))) 

fig
```
We have plotted the locations of earthquakes on the map, and it is evident that there is a substantial amount of data. Earthquakes primarily occur in regions of Canada that are in proximity to the United States and Greenland.
British Columbia, New Brunswick, and Yukon experience earthquakes more frequently than other provinces. British Columbia and Yukon are located on the Pacific Plate boundary, where the Pacific Plate is subducting beneath the North American Plate. This subduction can lead to significant seismic activity. New Brunswick is near the St. Lawrence rift system, which is associated with fault lines and seismic activity.
